---
title: "Rainfall_description"
author: "hui.shi@noaa.gov"
date: "2024-01-03"
output: md_document
preserve_yaml: FALSE
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(markdown)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2023

# Identify the variable we're focused on
var <- 'Rainfall'
```

```{r, include = FALSE, global = TRUE}
# Access indicator data
indicator_data <- read_csv(here('Indicator_Dashboard','Data', paste('Dashboard_Data_', RptYr, '.csv', sep = "")))

# And just use our variable of interest
indicator_data <- indicator_data |> filter(ID == var)
```

```{r, include = FALSE}
# Values that are included in the text

# Annual mean for report year
ann_mean_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Increase over time from linear model
if (!is.na(indicator_data$Value_lm[1])) {
  delta_rain_lm <- indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
    indicator_data$Value_lm[1]
  #delta_rain_pct = (indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
                      #indicator_data$Value_lm[1]) / indicator_data$Value_lm[1] * 100
}

# Increase in anomaly over time from linear model
if (!is.na(indicator_data$Anom_lm[1])) {
  delta_rain_anom_lm <- indicator_data$Anom_lm[length(indicator_data$Anom_lm)] - 
    indicator_data$Anom_lm[1]
}

# Monthly min & max for report year
monthly_min_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# Monthly min & max for previous years
monthly_min_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# long-term trend
time=1:length(indicator_data$Value)
m1=lm(indicator_data$Value~time)
trend <- coef(m1)[2] *12

```

## Rainfall
__About the Indicator__  
Rainfall may have substantive effects on the nearshore environment and is a potentially important co-variate with the landings of particular stocks.

__Status of the Indicator__  
The rainfall over the Main Hawaiian Islands was `r signif(ann_mean_RptYr,3)` mm day^-^^1^ in `r RptYr`. Monthly rainfall ranged from `r signif(monthly_min_RptYr,3)`–`r signif(monthly_max_RptYr,3)` mm day^-^^1^ during the year, within the range of values observed over the period of record (`r signif(monthly_min_PrevYrs,3)`–`r signif(monthly_max_PrevYrs,3)`  mm day^-^^1^). Over the period of record, there has been weakly significant decline of `r abs(signif(trend, 3))` mm day^-^^1^ yr^-^^1^ in rainfall. Over the time series, rainfall has declined by `r abs(signif(delta_rain_lm,2))`  mm day^-^^1^. The monthly anomaly has declined by `r abs(signif(delta_rain_anom_lm,2))`  mm day^-^^1^. Average rainfall was below average in the Main Hawaiian Islands domain.

__Indicator Data__  
Description: The CPC Merged Analysis of Precipitation (CMAP) is a technique which produces pentad and monthly analyses of global precipitation in which observations from rain gauges are merged with precipitation estimates from several satellite-based algorithms, such as infrared and microwave (NOAA 2002). The analyses are on a 2.5 x 2.5-degree latitude/longitude grid and extend back to 1979. CMAP Precipitation data are provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at https://www.esrl.noaa.gov/psd/. The data are comparable (but should not be confused with) similarly combined analyses by the Global Precipitation Climatology Project described in Huffman et al. (1997).
It is important to note that the input data sources to make these analyses are not constant throughout the period of record. For example, SSM/I (passive microwave - scattering and emission) data became available in July 1987; prior to that the only microwave-derived estimates available are from the MSU algorithm (Spencer, 1993) which is emission-based thus precipitation estimates are available only over oceanic areas. Furthermore, high temporal resolution IR data from geostationary satellites (every 3-hr) became available during 1986; prior to that, estimates from the OPI technique (Xie and Arkin 1997) are used based on OLR from orbiting satellites.
The merging technique is thoroughly described in Xie and Arkin (1997). Briefly, the methodology is a two-step process. First, the random error is reduced by linearly combining the satellite estimates using the maximum likelihood method, in which case the linear combination coefficients are inversely proportional to the square of the local random error of the individual data sources. Over global land areas the random error is defined for each time period and grid location by comparing the data source with the rain gauge analysis over the surrounding area. Over oceans, the random error is defined by comparing the data sources with the rain gauge observations over the Pacific atolls. Bias is reduced when the data sources are blended in the second step using the blending technique of Reynolds (1988). 

Timeframe: Monthly

Region: Main Hawaiian Island Grid (18.5° – 22.5°N, 161° – 154°W)

Measurement Platform: In-situ station gauges and satellite data.

Data available at: <http://apdrc.soest.hawaii.edu/erddap/griddap/hawaii_soest_bb7b_ce6a_f808.graph>.  

Sourced from: NOAA ESRL (2024).

__References__  
NOAA ESRL (2024): <http://apdrc.soest.hawaii.edu/erddap/griddap/hawaii_soest_bb7b_ce6a_f808.graph> Accessed: 4 April 2024. 

Xie, P., and P.A. Arkin (1997): Global precipitation: A 17-year monthly analysis based on gauge observations, satellite estimates, and numerical model outputs. Bull. Amer. Meteor. Soc., 78, 2539 - 2558.<https://journals.ametsoc.org/view/journals/bams/78/11/1520-0477_1997_078_2539_gpayma_2_0_co_2.xml>

nmfspalette: Stawitz C (2023): nmfspalette: A Color Palette for NOAA Fisheries. R package version 3.0.0.000. <https://nmfs-fish-tools.github.io/nmfspalette/>